---
title       : Gaussian Processes
author      : Bartek Skorulski
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---


```{r libraries, echo=FALSE, messages=FALSE}
require(knitr)
require(MASS)

require(data.table)
require(reshape2)

require(ggplot2)
```


## Definitions

_Gaussian process_ is a stochastic process where any finite number of
random variables have a joint Gaussian distribution.

A Gaussian process is completely specified by its _mean function_
$m(x)$ and _covariance function_\(k(x,x')\). They are defined by

\[m(x) = \mathcal{E}[f(x)],\]

\[ k(x,x') = E\big[\big(f(x) − m(x))(f(x') − m(x')\big)\big],\]

and will write the Gaussian process as \( f(x) \sim GP(m(x), k(x,
x'))\).


--- .class #id 

## Squared exponential

\[ cov(f(x_p), f (x_q)) = k(x_p , x_q) =
    \exp\Big(−\frac{1}{2} |x_p − x_q |\Big)^2.\]


```{r SigmaDef, echo=FALSE, messages=FALSE}

squared.exp <- function(x,y,l=1) {
  exp(-0.5* (norm(x-y, type="2")/l)^2)
}

Sigma <- function(x1, x2, cov.fun=squared.exp) {
  n1 <- length(x1)
  n2 <- length(x2)
  S <- matrix(rep(0, n1*n2), nrow=n1)
  for (i in 1:n1) {
    for (j in 1:n2) {
      S[i,j] <- cov.fun(x1[i],x2[j]) }}
  S
}

melted.simulations <- function(x.star, sim.number, ms, S) {
  sims <- cbind(data.table(x=x.star), t(mvrnorm(sim.number, ms, S)))
  melt(sims, id=1)
}

gg.simulations <- function(melted.sims) {
  ggplot() +
      geom_line(data = melted.sims, aes(x=x,y=value,
                    group=factor(variable)), alpha=.6) +
          theme_bw()

}


```

```{r ,echo=FALSE }
x.star <- seq(-5,5,0.1)
sim.number <- 5
S <- Sigma(x.star,x.star)
msims <- melted.simulations(x.star, sim.number,
                            rep(0, length(x.star)), S)
gg.simulations(msims)
```

---


## Brownian Motion

\[ cov(f(x_p), f (x_q)) = k(x_p , x_q) = \min(x_p, x_q).\]


```{r, echo=FALSE}
brow <- function(x1,x2) {
  min(x1,x2)
}

x.star <- seq(0, 5, 0.005)
S <- Sigma(x.star, x.star, brow)
ms <- rep(0, length(x.star))
msims <- melted.simulations(x.star, sim.number, ms, S)
gg.simulations(msims)

bm <- data.table(t(mvrnorm(2, rep(0, length(x.star)), S)))
ggplot(bm, aes(V1,V2))+
    geom_path() + theme_bw()
```

---

## Predictions

\[\overline{y}^* = K^*(\lambda^2\cdot I+K)^{-1}\cdot y \]

\[C = K^{**}-K^*\cdot(\lambda^2\cdot I+K)^{-1}\cdot(K^*)^t \]

```{r, echo=FALSE}

y.predict <- function(x, y, x.star, cov.fun, sigma=0) {
  K <- Sigma(x,x)
  I <- diag(length(x))
  K.star <- Sigma(x.star, x)
  K.star %*% solve(K+sigma^2*I) %*% y
}

cov.predict <- function(x, y, x.star, cov.fun, sigma=0) {
  K <- Sigma(x,x)
  I <- diag(length(x))
  K.star <- Sigma(x.star, x)
  K.star.star <- Sigma(x.star, x.star)
  K.star.star - K.star %*% solve(K+sigma^2*I) %*% t(K.star)
}

gg.simulations2 <- function(x, y, x.star, cov.fun, sim.number, sigma=0) {
  observations <- data.table(x, y)
  y.star.bar <- y.predict(x, y, x.star, cov.fun, sigma)[,1]
  predictors <- data.table(x=x.star, y=y.star.bar)
  S <- cov.predict(x, y, x.star, cov.fun, sigma)
  ms <- y.star.bar
  msims <- melted.simulations(x.star, sim.number, ms, S)
  d <- abs(diag(S))
  predictors$sigma <- sqrt(d)
  
  ggplot() +
      geom_line(data = msims, aes(x=x, y=value,
                    group=factor(variable)), alpha=.3, colour="green") +
          geom_point(data=observations,
                     aes(x,y), size=3, colour="red")+
              geom_line(data=predictors, aes(x, y), colour='blue') +
                  geom_ribbon(data=predictors,
                              aes(x, ymax=y+2*sigma,
                                  ymin=y-2*sigma), alpha=0.1)+
                      theme_bw()
}    

x <- c(-4,-3,-1,0,2)
y <- c(-2,0, 1, 2, -1)
x.star <- seq(-5,5,0.1)
sim.number <- 20
cov.fun <- squared.exp;

gg.simulations2(x, y, x.star, cov.fun, sim.number)
```
